UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-11-16 ms/main.tex
removed comments, changed some HRS Hispanic image and HRS_1KGP combined projection, rewrote abstract, minor ce

\begin{abstract}
%Max 250
Background: Population structure in genetic data depends on complex demographic processes including geographic isolation, genetic drift, migration, and admixture. Together with technical artifacts, population structure is a prominent confounder of genomic studies. Identifying such patterns is therefore central to the genomic enterprise. Whereas many methods can identify specific types of population structure, few are able to provide simple representations of genomic diversity across a range of scales. We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP).
\\
\\
Results: We demonstrate using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank that projections using PCA-UMAP effectively cluster individuals who are genetically closely related while placing them in a global continuum of genetic variation. These projections reveal non-trivial population groupings, reflect ethnicity and geography on fine-scale levels, and uncover patterns in the distributions of a variety of phenotypes.
\\
\\Conclusion: These projections succinctly illustrate population structure in large cohorts and capture relationships on local and global scales, establishing PCA-UMAP as a general-purpose approach to exploratory analysis in genomics.
\end{abstract}

@@

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. %Relatedness patterns are shaped by geographic isolation, founder effects, migration patterns, and admixture. Understanding such patterns is therefore useful to understand the genetic history of individuals and populations. It is also important to interpret genomic association studies, as unaccounted-for relatedness patterns can generate spurious correlations between genotypes and phenotypes. 
If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

@@

We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}.  
%A common practice in dimensional reduction is to first apply PCA before performing nonlinear dimensional reduction. We then then feed the top principal components into UMAP and tSNE to generate a two-dimensional representation of the data. After the projections are generated, we create dimensional images colored by variables such as ethnicity (or a proxy for ethnicity), admixture, geography, and phenotype.

\begin{figure}
\centering
   %\includegraphics[width=0.6\linewidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
   \caption{UMAP on the first 10 principal components of HRS data. Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}. Individuals with intermediate admixture ratios sometimes fall between clusters of Black, White, and Hispanic individuals \adpcomment{(alternate projection shown in figure~\ref{fig:supp_hrs_admix_alt}).} }
    %\caption{UMAP on the first 10 principal components of HRS data. coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. Individuals with intermediate admixture ratios sometimes fall between three clusters of Black, White, and Hispanic individuals. This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}}
    \label{fig:umap_hrs_admix}
\end{figure}

\subsection*{Results}
\begin{figure}
\centering
   %\includegraphics[width=0.6\linewidth]{images/HRS_HISP_UMAP_PC7_NC2_NN15_MD05_201844193148.pdf}
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_highlighted.pdf}
    \caption{The top $7$ principal components of the Hispanic population of the HRS projected by UMAP, colored by region of birth. The highlighted region consists almost entirely of individuals who were born in the Mountain region of the United States. This region contains New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same figure colored according to estimated continental ancestry proportion. Figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} shows that populations from the 1KGP do not map to the cluster when projected onto the UMAP embedding.}
    \label{fig:hrs_hisp}
\end{figure}

@@

\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying PCA-UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region. One separate cluster, highlighted in figure~\ref{fig:hrs_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding (figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}). The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    %\includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_2018330153123_born.pdf}
    \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_born.pdf}
    \end{subfigure}
    \caption{UMAP on the top 10 principal components of the HRS dataset, colored by Census Bureau birth region. There is no obvious pattern in the clusters of majority "White Not Hispanic" individuals.}
    \label{fig:supp_hrs_born}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_custom_label.pdf}
    \end{subfigure}
    \caption{\adpcomment{This is also new} UMAP on the top 10 principal components of the HRS data, with 1KGP data projected onto the embedding. Individuals from the HRS are coloured grey and largely transparent. British (GBR) and other European (CEU) individuals are scattered throughout the "White Not Hispanic" clusters. Finns (FIN) form clear groupings. Spanish (IBS) and Italian (TSI) individuals cluster near the Hispanic grouping. There is a clear gradient in the Hispanic cluster formed of Puerto Ricans (PUR), Colombians (CLM), Mexicans (MXL), and Peruvians (PEL). Populations with African ancestry (AFR) appear with Black individuals. East Asian (EAS) populations comprising Chinese, Kinh, and Japanese individuals cluster together with what appears in figure~\ref{fig:umap_hrs_admix} as a population of mostly Asian ancestry. South Asian (SAS) populations with Indian, Pakistani, and Sri Lankan ancestry cluster in a separate area. One "White Not Hispanic" cluster at the bottom does not cluster with any 1KGP populations.}
    \label{fig:supp_hrs_1kgp_projected}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
    \end{subfigure}
    \caption{UMAP on the top 10 principal components of the combined HRS and 1KGP datasets. The colors match those used in the separate 1KGP and HRS plots. The large blue clusters are "White Not Hispanic" individuals. Scattered brown dots within these clusters are Utah residents with Northern/Western European ancestry, and British in England and Scotland ("CEU" and "GBR", respectively, in the 1KGP datasets). A cluster of Finnish individuals consistently shows up and is labelled "FIN". Toscani individuals regularly form clusters and are labelled "TSI". However, their locations relative to the other clusters of "White Not Hispanic" are not consistent, sometimes appearing next to clusters of Spanish individuals and other times not. One cluster of "White Not Hispanic" individuals is highlighted in a box --- unlike the other clusters, CEU and GBR individuals are not projected here.}
    \label{fig:supp_hrs_1000g}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    %\includegraphics[width=\textwidth]{images/HRS_HISP_UMAP_PC7_NN15_MD05_admix.pdf}
    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \end{subfigure}
    \caption{UMAP of the first $7$ principal components of the Hispanic population of the HRS, colored by birth region, with selected 1KGP populations projected onto the UMAP embedding. The highlighted area consists almost entirely of individuals born in the Mountain region of the census (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). The six populations do not fall onto the highlighted cluster.  CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy.}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
    \end{subfigure}
    \caption{UMAP on the first $10$ principal components of HRS data, coloring individuals by estimated admixture from three ancestral populations. Alternate projection to demonstrate how individuals may fall between clusters.}
    \label{fig:supp_hrs_admix_alt}
\end{figure*}

>>>>

\begin{abstract}
%Max 250
Background: Population structure in genetic data depends on complex demographic processes including geographic isolation, genetic drift, migration, and admixture. Existing data exploration methods struggle with unbalanced cohort ancestries, reconciliation of global versus local relationships, and the sheer size of modern datasets. We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
\\
\\
Results: We demonstrate using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank that projections using PCA-UMAP effectively cluster individuals who are genetically closely related while placing them in a global continuum of genetic variation. These projections quickly reveal subpopulations within the American Hispanic population, deep relationships between geography and genotype in the UK population, and stark phenotypic variation between groups in the UK Biobank. 
\\
\\
Conclusion: These projections succinctly illustrate population structure in large cohorts and capture relationships on local and global scales, establishing PCA-UMAP as a general-purpose approach to exploratory analysis in genomics.
\end{abstract}

@@

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

@@

We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}.

\begin{figure}
\centering
   %\includegraphics[width=0.6\linewidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
   \caption{UMAP on the first 10 principal components of HRS data. Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
    \label{fig:umap_hrs_admix}
\end{figure}

%\subsection*{Results}
%\begin{figure}
%\centering
%   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_highlighted.pdf}
%    \caption{The top $7$ principal components of the Hispanic population of the HRS projected by UMAP, colored by region of birth. The highlighted region consists almost entirely of individuals who were born in the Mountain region of the United States. This region contains New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same figure colored according to estimated continental ancestry proportion. Figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} shows that populations from the 1KGP do not map to the cluster when projected onto the UMAP embedding.}
%    \label{fig:hrs_hisp}
%\end{figure}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.6\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \end{subfigure}
    \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six relevant 1KGP populations are presented: CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
\end{figure*}

@@

\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying PCA-UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region. One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_born.pdf}
    \end{subfigure}
    \caption{UMAP on the top 10 principal components of the HRS dataset, colored by Census Bureau birth region. There is no obvious pattern in the clusters of majority "White Not Hispanic" individuals.}
    \label{fig:supp_hrs_born}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_custom_label.pdf}
    \end{subfigure}
    \caption{UMAP on the top 10 principal components of the HRS data, with 1KGP data projected onto the embedding. Individuals from the HRS are grey. British (GBR) and other European (CEU) individuals are scattered throughout the "White Not Hispanic" clusters. Finns (FIN) form clear groupings. Spanish (IBS) and Italian (TSI) individuals cluster near the Hispanic grouping. There are sub-groups in the Hispanic cluster formed of Puerto Ricans (PUR), Colombians (CLM), Mexicans (MXL), and Peruvians (PEL). Populations with African ancestry (AFR) appear with Black individuals. East Asian (EAS) populations comprising Chinese, Kinh, and Japanese individuals cluster together with what appears in figure~\ref{fig:umap_hrs_admix} as a population of mostly Asian ancestry. South Asian (SAS) populations with Indian, Pakistani, and Sri Lankan ancestry cluster in a separate area. One "White Not Hispanic" cluster at the bottom does not cluster with any 1KGP populations.}
    \label{fig:supp_hrs_1kgp_projected}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
    \end{subfigure}
    \caption{UMAP projection of the top 10 principal components of the combined HRS and 1KGP datasets. One cluster (in the box) does not group with any of the 1KGP populations. A cluster of Finnish (FIN) individuals consistently appears in the "White Not Hispanic" (WNH) group. Groups of Central and South American populations from the 1KGP (CLM, Colombian; MXL, Mexican; PEL, Peruvian; PUR, Puerto Rican) form nearby or within the HRS Hispanic cluster (HIS). Iberian individuals (IBS) cluster near the Hispanic population. Toscani individuals (TSI) form some small clusters and sometimes appear near the Iberian and Hispanic populations. Individuals with British/Scottish (GBR) or Northern/Western European ancestry (CEU) are scattered throughout the WNH clusters. Individuals with African ancestry from the 1KGP group with Black Americans from the HRS (AFR). Similar population groupings occur with South Asian (SAS) and East Asian (EAS) individuals.}
    \label{fig:supp_hrs_1000g}
\end{figure*}

@@

%\begin{figure*}
%    \centering
%    \begin{subfigure}{\textwidth}
%    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
%    \end{subfigure}
%    \caption{UMAP of the first $7$ principal components of the Hispanic population of the HRS, colored by birth region, with selected 1KGP populations projected onto the UMAP embedding. The highlighted area consists almost entirely of individuals born in the Mountain region of the census (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). The six populations do not fall onto the highlighted cluster.  CLM, Colombian in Medellin, Colombia;
%    IBS, Iberian in Spain;
%    MXL, Mexican in Los Angeles, California;
%    PEL, Peruvian;
%    PUR, Puerto Rican;
%    TSI, Tuscani in Italy.}
%    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
%\end{figure*}

@@

%\begin{figure*}
%    \centering
%    \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
%    \end{subfigure}
%    \caption{UMAP on the first $10$ principal components of HRS data, coloring individuals by estimated admixture from three ancestral populations. Alternate projection to demonstrate how individuals may fall between clusters.}
%    \label{fig:supp_hrs_admix_alt}
%\end{figure*}
